---
title: "Coordinating on-demand service supply and demand paper notes"
author: "Adam Spannbauer"
date: "`r Sys.Date()`"
output: 
  html_document:
    mathjax: "default"
    code_folding: hide
    df_print: paged
    theme: flatly
    toc: true
    toc_float: true
    toc_depth: 4
---

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(triangle)
library(patchwork)

plot_shaded_normal <- function(mean, sd, p = 0.5) {
  shade_up_to_x <- qnorm(p, mean = mean, sd = sd)

  data_normal <- data.frame(
    x = seq(mean - 4 * sd, mean + 4 * sd, length.out = 300)
  ) |>
    mutate(y = dnorm(x, mean = mean, sd = sd))

  shaded_data_normal <- data_normal |>
    filter(x <= shade_up_to_x)

  ggplot(data_normal, aes(x = x, y = y)) +
    geom_line() +
    geom_area(
      data = shaded_data_normal,
      aes(x = x, y = y),
      fill = "#ff8800",
      alpha = 0.3
    ) +
    geom_vline(xintercept = shade_up_to_x)
}

plot_shaded_triangle <- function(a, b, c = (a + b) / 2, p = 0.5) {
  shade_up_to_x <- triangle::qtriangle(p, a = a, b = b, c = c)

  data_tri <- data.frame(
    x = seq(a, b, length.out = 300)
  ) |>
    mutate(y = triangle::dtriangle(x, a = a, b = b, c = c))

  shaded_data_tri <- data_tri |>
    filter(x <= shade_up_to_x)

  ggplot(data_tri, aes(x = x, y = y)) +
    geom_line() +
    geom_area(
      data = shaded_data_tri,
      aes(x = x, y = y),
      fill = "#ff8800",
      alpha = 0.3
    ) +
    geom_vline(xintercept = shade_up_to_x)
}

```

Reading paper:

Bai, J., So, K. C., Tang, C. S., Chen, X., & Wang, H.
(2019).
Coordinating supply and demand on an on-demand service platform with impatient customers.
Manufacturing & Service Operations Management, 21(3), 556-570.
<https://doi.org/10.1287/msom.2018.0707>

------------------------------------------------------------------------

# Notes

## General overview and goals

-   Customers arrive randomly
-   Each service consists of random amount of service units (e.g. travel distance)
    -   Assumed any service can be rendered by any agent (maybe not true in reality where drivers cancel trips based on distance)
-   Fixed price rate $p$ per service unit (e.g. dollars per kg)
-   Fixed wage rate $w$ per service unit
-   Payout ratio: $\frac{w}{p}$

No focus on surge pricing for ease of modeling and cites customer dissatisfaction if price jumps in short time span.

Assumed $p$ and $w$ are known to customers and agents in advance.

Goal of service platform is to set $p^*$ and $w^*$ to maximize average profit.

## Realized customer rate $\lambda$ and price rate $p$

### Setup

-   Let $\lambda$ be the customer request rate with max amount of service in a time period is $\bar{\lambda}$

-   Customer value per service unit is $v$ (heterogeneous with CDF $F(\cdot)$)

-   For a customer with valuation $v$ and service request of $D$ units the surplus is $(v - p)D$

    -   $D$ is assumed to be independent of customer-type $v$

    -   expected value of service units will be modeled as $d = E(D)$

-   Utility function of a customer can be modeled as $U(v) = (v - p)d - cW_q$

    -   $v$ is value per service unit

    -   $p$ is price per service unit

    -   $d$ is expected value of service units

    -   $c$ is cost of waiting

    -   $W_q$ is expected wait time for service

-   Realized customer request rate $\lambda$ is the percent of $\bar{\lambda}$ of customers who have $U(v) \ge 0$ for a given $p$

    -   aka $\lambda = \text{Prob} \left\{ U(v) \ge 0 \right\} \cdot \bar{\lambda} \to \lambda = \text{Prob} \left\{ v \ge p + \frac{c}{d} W_q \right\} \cdot \bar{\lambda}$

    -   steps

        -   $(v-p)d-cW_q \ge 0$

        -   $vd-pd-cW_q \ge 0$

        -   $vd \ge pd-cW_q$

        -   $v \ge p - \frac{c}{d}W_q$

-   This probability is labeled $s$ (i.e. $s = \text{Prob} \left\{ v \ge p + \frac{c}{d} W_q \right\}$ and $\lambda = s \bar{\lambda}$) and represents the desired service rate.
    We can manipulate $p$ to decide what percentage of $\bar{\lambda}$ to target

-   Since $v \sim F(\cdot)$ then the price rate satisfies $p = F^{-1} \left( 1 - \frac{\lambda}{\bar{\lambda}} \right) - \frac{c}{d} W_q$ (note this could be written $p = F^{-1} \left( 1 - s \right) - \frac{c}{d} W_q$ to show service rate more directly)

    -   Steps

        -   $\lambda = \text{Prob} \left\{ v \ge p + \frac{c}{d} W_q \right\} \cdot \bar{\lambda}$

        -   $\lambda = \left[ 1 - F\left( p + \frac{c}{d} W_q \right) \right] \cdot \bar{\lambda}$

        -   $\frac{\lambda}{\bar{\lambda}} = 1 - F\left( p + \frac{c}{d} W_q \right)$

        -   $F\left( p + \frac{c}{d} W_q \right) = 1 - \frac{\lambda}{\bar{\lambda}}$

        -   $p + \frac{c}{d} W_q = F^{-1}\left(1 - \frac{\lambda}{\bar{\lambda}}\right)$

        -   $p = F^{-1}\left(1 - \frac{\lambda}{\bar{\lambda}}\right) - \frac{c}{d} W_q$ or $p = F^{-1}\left(1 - s\right) - \frac{c}{d} W_q$

    -   Holding everything else constant

        -   Price rate decreases as wait time ( $W_q$ ) increases

        -   Price rate decreases as unit wait cost ( $c$ ) increases

        -   Price rate increases as number of service units ( $d$ ) increases

### Final results

For a given service rate ( $s$ ), wait time ( $W_q$ ), expected service units ( $d$ ), cost of wait per unit time ( $c$ ), & distribution of $v$ ( $v \sim F(\cdot)$ ) we can prescribe a price to obtain $s$:

$$
p = F^{-1}\left(1 - s\right) - \frac{c}{d} W_q
$$

Customers assign different values ( $v$ ) for the service per unit.
$v \sim F( \cdot )$.
Exploring normal and triangle distributions for ease.
Likely a right skewed distribution in practice, but perhaps those out in the tail of the a heavy right skewed dist are out of the customer base for these on demand services.

```{r}
# Price rate as a function of service level
price_rate <- function(s, c, d, Wq, Finv) {
  Finv(1 - s) - (c / d) * Wq
}
```

#### How price varies with individual params

Holding all other factors constant and just manipulating the input shown on x axis.

Service level $s$ is not plotted, but as desired service level is increases, the prescribed price decreases (make cheaper to get more of market; make more expensive to out price parts of market).

These relationships are the similar in behavior regardless of choice of value distribution family ( $F(\cdot)$ ).

```{r  fig.height=3, fig.align="center"}
ex_range <- 0:10

s <- 0.5
c <- 2
d <- 2
Wq <- 2

# value dist (normal) params
mu <- 5
sd <- 1
Finv <- \(p) qnorm(p = p, mean = mu, sd = sd)

ex_c <- price_rate(s, ex_range, d, Wq, Finv)
ex_d <- price_rate(s, c, ex_range, Wq, Finv)
ex_Wq <- price_rate(s, c, d, ex_range, Finv)

func_behavior_df <- data.frame(
  ex_range = ex_range,
  c = ex_c,
  d = ex_d,
  Wq = ex_Wq
)

func_behavior_df |>
  pivot_longer(-ex_range) |>
  mutate(name = ifelse(name == "c", "Cost of waiting (c)", name)) |>
  mutate(name = ifelse(name == "d", "Average service units (d)", name)) |>
  mutate(name = ifelse(name == "Wq", "Wait time (Wq)", name)) |>
  ggplot(aes(x = ex_range, y = value)) +
  geom_line() +
  geom_hline(yintercept = 0, linetype = "dotted") +
  facet_wrap(~name) +
  scale_x_continuous(breaks = seq(0, 10, 2)) +
  labs(
    y = "price", x = "",
    title = bquote("Behavior of price with inputs; " ~
      p == F^
        {
          -1
        } * (1 - s) - frac(c, d) * W[q]),
    subtitle = "Holding all other factors constant & service level (s) = 0.5"
  )
```

#### Triangle distribution of value $v \sim F_{triangle}(\cdot)$

```{r fig.height=3, fig.align="center"}
c <- 2
d <- 2
Wq <- 2

# value dist (triangle) params
a <- 1
b <- 10
c <- 2
Finv <- \(p) qtriangle(p = p, a = a, b = b, c = c)

s <- 0.2
p1 <- plot_shaded_triangle(a = a, b = b, c = c, p = s) +
  labs(
    title = sprintf(
      "For service level %.0f%% -> price = %.2f ",
      s * 100, price_rate(s, c, d, Wq, Finv)
    ),
    subtitle = sprintf("Given c = %.1f; d = %.1f, Wq = %.1f", c, d, Wq),
    x = "v",
    y = ""
  ) +
  theme(
    plot.title = element_text(size = 10),
    plot.subtitle = element_text(size = 8)
  )


s <- 0.6
p2 <- plot_shaded_triangle(a = a, b = b, c = c, p = s) +
  labs(
    title = sprintf(
      "For service level %.0f%% -> price = %.2f ",
      s * 100, price_rate(s, c, d, Wq, Finv)
    ),
    subtitle = sprintf("Given c = %.1f; d = %.1f, Wq = %.1f", c, d, Wq),
    x = "v",
    y = ""
  ) +
  theme(
    plot.title = element_text(size = 10),
    plot.subtitle = element_text(size = 8)
  )


p1 + p2
```

#### Normal distribution of value $v \sim F_{normal}(\cdot)$

```{r fig.height=3, fig.align="center"}
c <- 2
d <- 2
Wq <- 2

# value dist (normal) params
mu <- 5
sd <- 1
Finv <- \(p) qnorm(p = p, mean = mu, sd = sd)

s <- 0.3
p1 <- plot_shaded_normal(mean = mu, sd = sd, p = s) +
  labs(
    title = sprintf(
      "For service level %.0f%% -> price = %.2f ",
      s * 100, price_rate(s, c, d, Wq, Finv)
    ),
    subtitle = sprintf("Given c = %.1f; d = %.1f, Wq = %.1f", c, d, Wq),
    x = "v",
    y = ""
  ) +
  theme(
    plot.title = element_text(size = 10), # Shrink title font size
    plot.subtitle = element_text(size = 8) # Shrink subtitle font size
  )

s <- 0.8
p2 <- plot_shaded_normal(mean = mu, sd = sd, p = s) +
  labs(
    title = sprintf(
      "For service level %.0f%% -> price = %.2f ",
      s * 100, price_rate(s, c, d, Wq, Finv)
    ),
    subtitle = sprintf("Given c = %.1f; d = %.1f, Wq = %.1f", c, d, Wq),
    x = "v",
    y = ""
  ) +
  theme(
    plot.title = element_text(size = 10),
    plot.subtitle = element_text(size = 8)
  )

p1 + p2
```

## Realized providers $k$ and wage rate $w$

-   $K$ - maximum number of potential providers
-   $k$ - realized number of providers for a given price $p$ and wage rate $w$ (wage as a percentage - $\alpha$ - of $p$)
    -   $k \le K$
-   $\mu$ - average service speed; so $\frac{\mu}{d}$ is service rate (i.e. average number of customers served per hour; $d$ is average service units requested)
-   Given realized customers ( $\lambda$ ) and realized providers ( $k$ ), the utilization is $\frac{\lambda}{k \cdot (\mu / d)} = \frac{\lambda d}{k \mu}$
    -   Aka $\rho = \frac{\lambda d}{k \mu}$ for $M/M/k$ queue
    -   Similar to typical queueing $\rho = \frac{\lambda}{\mu}$, but incorporating $k$ providers and $d$ average service units
-   Wage per unit time of a single provider is $w\mu$
-   Average earning rate is $w\frac{\lambda d}{k}$
    -   that is: single provider earning $w \mu$ times utilization $\frac{\lambda d}{k \mu}$
    -   $w \mu \cdot \frac{\lambda d}{k \mu} = w\frac{\lambda d}{k}$
-   Each provider has a "reservation earning rate" $r$ that has CDF $G(\cdot)$
    -   i.e. they won't average earning rate $\ge$ than reservation rate (aka $w\frac{\lambda d}{k} \ge r$ )
-   $\beta$ will denote proportion of providers partipating
    -   $\beta = \text{Prob} \left\{ r \le w(\lambda d / k) \right\} = G \left( w(\lambda d / k) \right)$
    -   Also: $G^{-1}(\beta) = w \frac{\lambda d}{k}$
    -   This means realized providers $k$ is $k = \beta K$
-   We can express wage rate $w$ as a function of participating providers $k$
    -   $w = G^{-1}(\beta) \frac{k}{\lambda d} = G^{-1}(\frac{k}{K}) \frac{k}{\lambda d}$

## Objective: max profit

-   Profit ( $\pi$ ) is $\pi = \lambda (p - w) d$
    -   $\lambda$ - customer requests
    -   $p$ - price rate per service unit
    -   $w$ - wage rate per service unit
    -   $d$ - average service units
-   We can plug in the definitions of $p$ (in terms of service rate - $\lambda$) and $w$ (in terms of participating provider rate - $k$)
    -   $\pi(k, \lambda) = \lambda d \left[ \left( F^{-1}\left(1 - \frac{\lambda}{\bar{\lambda}}\right) - \frac{c}{d} W_q \right) - \left( G^{-1}(\frac{k}{K}) \frac{k}{\lambda d} \right) \right]$
    -   Or written with $s$ to denote request rate and $\beta$ to denote provider participation rate: $\max_{k, \lambda} \pi(k, \lambda) \equiv \lambda d \left[ \left( F^{-1}\left(1 - s\right) - \frac{c}{d} W_q \right) - \left( G^{-1}(\beta) \frac{k}{\lambda d} \right) \right]$
    -   Subject to $\frac{\lambda d}{k \mu} < 1$ (i.e. the requests don't outpace the service)
        -   Note: maybe not a realistic constraint for peak event times, but maybe surge pricing and wait times can diminish request rate enough to hold
-   We'll use $\pi(k, \lambda)$ to find optimal supply $k^*$ and optimal demand $\lambda^*$, and in turn we can use $k^*$ and $\lambda^*$ to find optimal price $p^*$ and optimal wage rate $w^*$

## Base model

-   Wage rate ( $w$ ) will be a proportion ( $\alpha$ ) of price rate ( $p$ ) - $w = \alpha p$

-   Waiting time $W_q$ will be based on $M/M/k$ queue with arrival rate $\lambda$ and service rate $\frac{\mu}{d}$

    -   $W_q = \frac{1}{1 + \left( \frac{k! (1 - \rho)}{k^k \rho^k} \right)
        \sum_{i=0}^{k-1} \frac{k^i \rho^i}{i!}} \left[ \frac{\rho}{\lambda (1 - \rho)} \right]$
    -   Where $\rho = \frac{\lambda d}{k \mu}$ is system utilization with $\rho < 1$

-   For simplicity in based model, distribution of $v$ (so far referred to as $F(\cdot)$ ) and $r$ (so far referred to as $G(\cdot)$ ) will both be modeled as uniform on range $[0, 1]$.
    This means we can drop out mention of them from current model.

-   Updated formulation with $W_q$ and uniform dists on $[0, 1]$

**STOPPING SPOT - To be continued...**

# Slides

<embed src="slides.pdf#view=Fit" type="application/pdf" width="100%" height="600px">
